---
title: "PCA Example (Iris)"
author: "Hayden Ginman"
date: "2023-04-24"
output: html_document
---
Install the requisite packages to easily perform PCA and load the required libraries
```{r}
options(repos = c(CRAN = "https://cran.rstudio.com"))
# install.packages(c("tidyverse", "FactoMineR", "factoextra", "ggcorrplot"))
library(tidyverse)
library(FactoMineR)
library(factoextra)
library(ggcorrplot)

```
1. Load the Iris dataset
```{r}
data("iris")

```
2. Briefly explore dataset
```{r}
head(iris)
summary(iris)

```

3. Normalize the dataset
```{r}
#Removes species column to isolate numerical values
iris_norm <- iris %>% select(-Species)
#Scales the data: centers the data by subtracting the mean of each variable from each observation, and then divide by the standard deviation of each variable, or by the square root of the variance. 
iris_scale <- scale(iris_norm)

```
4. Compute and visualize the covariance matrix
```{r}
#Covariance matrix displays the pairwise covariance between variables, which is a measure of how two variables vary together.
cov_matrix_scaled <- cov(iris_scale)

print(cov_matrix_scaled)
#Heatmap visualization makes it easier to see relationships between the variables
heatmap_plot <- ggcorrplot(cov_matrix_scaled, hc.order = TRUE, type = "lower",
                           lab = TRUE, lab_size = 3, tl.cex = 12, title = "Heatmap of Covariance Matrix")
print(heatmap_plot)

```




5. Calculate and display principal components
```{r}
#Base R function that performs PCA on a dataset
p_comp <- prcomp(iris_scale)
print(p_comp)

```

6. Perform PCA on the dataset
```{r}
#FactoMineR can also be used to perform PCA of the dataset
pca <- PCA(iris_scale, scale.unit = TRUE, graph = FALSE)

```
7. Print a summary of the PCA results
```{r}
#Allows us to see eigenvalues
summary(pca)
```
8. Visualize the PCA results
```{r}
#A visualization that helps us to determine the number of principal components to retain in a PCA
fviz_eig(pca, addlabels = TRUE, ylim = c(0, 100), ggtheme = theme_minimal())
#A visualization that helps us visualize the relationships between the variables and observations in a dataset in a low-dimensional space
fviz_pca_biplot(pca, label = "var", habillage = iris$Species,
                addEllipses = TRUE, ellipse.type = "confidence",
                palette = "jco", ggtheme = theme_minimal())

```








9. Transform the data
```{r}
#Calculate coefficients of the linear combinations of the original variables that are used to construct each principal component
loadings <- pca$var$coord[, 1:2]
transformed_data <- as.data.frame(iris_scale %*% loadings)
transformed_data$Species <- iris$Species
colnames(transformed_data) <- c("PC1", "PC2", "Species")

```
10. Visualize the transformed data
```{r}
#Scatterplot of transformed dataset that shows the relationship between the first 2 principal components
ggplot(transformed_data, aes(x = PC1, y = PC2, color = Species)) +
  geom_point(size = 3) +
  theme_minimal() +
  labs(title = "PCA of Iris Dataset",
       x = "Principal Component 1",
       y = "Principal Component 2",
       color = "Species")
```